preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution

### 2021-01-22 BMC_Genome_Biology_submission/grubert_manuscript_figures_tables/manuscript/manuscript.Rmd
minor edits for grammar and structure including new reference

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of the 3-dimensional (3D) structure of the human genome such as chromatin loops [@rao20143d], Topologically Associating Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At the kilobase scale, chromatin loops connect gene promoters with distal enhancers, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. At the megabase scale, TADs represent regions on the linear genome that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Identifying the precise location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify genomic coordinates of TADs (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised]. However, a key limitation of them is that they depend on Hi-C data *resolution*. Resolution refers to the size of genomic regions (bins) used to segment the linear genome [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which leads to increased uncertainty in the location of boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. While these low-resolution Hi-C datasets can be used to probe higher-order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TAD and loop boundaries [@fudenberg2016formation]. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth [@schmitt2016genome], obtaining high-resolution HI-C data and boundaries remains difficult.

@@

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin state (BroadHMM), histone modification (HM), and transcription factor binding site (TFBS) data to predict both low-resolution TAD and chromatin-loop boundaries. We introduce a systematic pipeline for building the optimal boundary region prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites improve prediction performance, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and partitioning around medoids (PAM) to detect genome annotation-guided boundary regions and points at a base-level resolution. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD*-predicted boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on https://bioconductor.org/packages/preciseTAD.

@@

<!--%%%% revised paragraph-->
We developed a machine learning (ML) framework for determining the optimal set of data level characteristics to predict boundary regions of Topologically Associating Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. For TAD prediction, we derived cell line specific boundaries at 5-100 kb resolutions using *Arrowhead* [@durand2016juicer] (Additional file 1: Arrowhead Script, Additional file 2: Table S1). For chromatin loop prediction, we downloaded loops derived by *Peakachu* [@salameh2019supervised] at 10 kb (Additional file 2: Table S1). Our models were trained on cell line-specific functional genomic annotation data (ChIP-seq peak regions) publicly available through ENCODE [@encode2004encode; @davis2018encyclopedia] (Additional file 4: Table S3). We used ChIA-PET derived loop boundaries [@grubert2020landscape] (Additional file 2: Table S1), referred hereafter as *Grubert* boundaries, as the gold standard when validating our models and extracting performance metrics (See Methods). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base-level resolution from cell line-specific genome annotation data.
<!--%%%-->

@@

Boundary regions were defined as genomic bins containing a called boundary ($Y = 1$), while non-boundary regions were defined as bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Additional file 3: Table S2). The number of non-boundary regions highly outnumbered boundary regions. Such a disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address the class imbalance, we evaluated the effect of three resampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority class (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

@@

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracy than the peak signal strength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions, with results consistent for K562 (Additional file 6: Figure S2A). Furthermore, 5 kb resolution genomic bins led to the optimal prediction for TAD boundary regions on both cell lines. Random forest models built on Peakachu-defined loop boundary regions yielded optimal prediction performance when using a distance-type predictor space, with SMOTE resampling only marginally outperforming RUS, for both cell lines (Additional file 6: Figure S2B, S2C). Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and chromatin loop boundary prediction.

@@

We sought to further optimize our boundary region prediction models by implementing recursive feature elimination to avoid overfitting and selecting only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Additional file 7: Figure S3B). However, given that we trained our models on chromosome-specific data, the most significant annotations varied for each holdout chromosome. To determine transcription factors most important for boundary prediction across all chromosomes, we clustered the predictive importance (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found four transcription factors, CTCF, RAD21, SMC3, and ZNF143, being consistently predictive of TAD and loop boundaries (Figure 4C; Additional file 7: Figure S3C). We optimized our model by only considering these top four TFBS when building the random forest model, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the known molecular drivers of the loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line using the model pre-trained on data from another cell line (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic annotation data from the GM12878 cell line resulted in an average AUC=<!--0.792--> <!--%%%--> 0.840<!--%%%--> (Figure 4D). Interestingly, when training on the K562 cell line and testing on GM12878, the average AUC increased slightly to <!--0.795--> <!--%%%--> 0.899<!--%%%-->. Likewise, the average performance of models trained using Peakachu boundaries and genomic annotation data from the GM12878 cell line was comparable to models trained on K562-specific Peakachu boundaries and genomic annotations (Avg. AUC=<!--0.881--> <!--%%%--> 0.923<!--%%%--> and <!--0.874--> <!--%%%--> 0.876<!--%%%-->, respectively). These results were consistent when comparing training/testing strategies on K562 with training on GM12878 and testing on K562 data (Additional file 7: Figure S3D). The average ROC curves were found to be within 1 standard deviation of each other, suggesting that a model trained on data from one cell line performs well when using the data from another cell line, allowing for the opportunity to predict boundaries for cell lines that do not currently have Hi-C data available.

@@

When trained using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb, respectively, the *preciseTAD* model predicted a total of 12,258 TAD and 15,707 chromatin loop boundaries in GM12878, as well as 9,603 TAD and 11,154 chromatin loop boundaries in K562 cell line (Additional file 9: Table S4). We reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Additional file 3: Table S2). This can be attributed to Arrowhead's inflation of called TADs at 5 kb, that, when visualized, often do not correspond to domains enriched in internal interactions (Figure 5A) and signal of known drivers of domain boundaries (Figure 5B). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Additional file 3: Table S2). This can be attributed to Peakachu's use of only CTCF sites to call boundaries, while *preciseTAD* leverages four known drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant than those detected from Hi-C data alone.

@@

Next, we evaluated the biological significance of *preciseTAD* boundary points. The signal of four known molecular drivers of 3D chromatin (CTCF, RAD21, SMC3, and ZNF143) colocalized more frequently around *preciseTAD*-predicted boundary points, as compared to Arrowhead-called TAD and Peakachu loop boundaries, respectively (<!--Figure 5C, 5D; Additional file 10: Figure S5A, S5B--> <!--%%%-->Figure 6A, 6B; Additional file 10: Figure S5A, S5B <!--%%%-->). <!--%%%-->Additionally, we found that when trained on cohesin-mediated chromatin loops, *preciseTAD*-predicted boundaries showed similar levels of enrichment to the set of ChIA-PET derived loop boundaries (Figure 6C; Additional file 10: Figure S5C).<!--%%%--> Surprised by the poor signal distribution around Arrowhead and Peakachu boundaries, we compared signals centered on boundaries called by Arrowhead, Peakachu, <!--%%%-->Grubert,<!--%%%--> and our SpectralTAD [@Cresswell:2020ab] TAD caller. We confirmed the poor signal distribution around Arrowhead-, Peakachu-, and SpectralTAD-called boundaries,in contrast to the well-performing Grubert-defined boundaries (<!--Additional file 10: Figure S5C, S5D--> <!--%%%-->Additional File 11: Figure S6<!--%%%-->). Signal enrichment heatmaps confirmed that *preciseTAD*-predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone<!--%%%-->, while showing similar levels of enrichment when compared to Grubert<!--%%%--> (<!--Additional file 11-12: Figure S6-S7--> <!--%%%-->Additional File 12-14: Figure S7-S9<!--%%%-->). *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBS (Wilcoxon p-value < 0.001 versus Arrowhead and Peakachu boundaries, respectively, <!--Figure 6A, Additional file 13: Figure S8-->  <!--%%%-->Figure 7A; Additional file 15: Figure S10<!--%%%-->). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.<!--%%%-->While *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBS compared to either Arrowhead and Peakachu boundaries alone, Grubert boundaries were found to be the most spatially proximal (Wilcoxon p-value < 0.001; Figure 7A; Additional file 15: Figure S10). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation when compared to a novel TAD-caller or loop prediction algorithm.<!--%%%-->

@@

Guided by the high predictive performance when training and testing on different cell lines (Figure 4D), we opted to evaluate whether models trained using Arrowhead/Peakachu<!--%%%-->/Grubert<!--%%%--> ground truth data in one cell line could be leveraged to predict boundaries using annotation data from another cell line. We evaluated two scenarios: 1) training on GM12878 and predicting boundaries on GM12878 (GM on GM) vs. training on K562 and predicting on GM12878 (K on GM), and 2) training on K562 and predicting boundaries on K562 (K on K) vs. training on GM12878 and predicting boundaries on K562 (GM on K). Using Arrowhead-trained models, 76% (J=0.701) and 81% (J=0.751) of predicted boundaries overlapped in both cross-cell-line prediction scenarios (<!--Additional file 15: Figure S10A, S10B--> <!--%%%-->Additional file 17: Figure S12<!--%%%-->). When using Peakachu-trained models, we observed 85% (J=0.705) and 88% (J=0.759) overlap (<!--Additional file 15: Figure S10C, S10D--> <!--%%%-->Additional file 18: Figure S13<!--%%%-->). <!--%%%-->Lastly, when using Grubert-trained models, we observed 86% (J=0.698) and 80% (J=0.0.660) overlap (Additional file 19: Figure S14).<!--%%%--> Furthermore, boundaries predicted on unseen annotation data exhibited a similar level of enrichment for CTCF, RAD21, SMC3, and ZNF143 as did those trained and predicted on the same cell line (Additional files 17-19: Figure S12-S14). These results indicate that *preciseTAD* pre-trained models can be successfully used to accurately predict both TAD and loop boundaries for cell lines lacking Hi-C data but for which genome annotation data is available.

@@

<!--%%%%revised paragraph-->
We present *preciseTAD*, a machine learning approach for precise prediction of TAD and chromatin loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on low-resolution domain boundaries obtained from domain calling tools, and high-resolution genomic annotations as the predictor space. *preciseTAD* predicts the probability of each base being a boundary, and identifies the precise location of boundary regions and the most likely boundary points. *preciseTAD* was benchmarked against multiple boundary calling tools including Arrowhead [@durand2016juicebox], an established TAD-caller, Peakachu [@salameh2019supervised], a recently published algorithm for predicting chromatin loops, and Grubert [@grubert2020landscape], a method for mapping cohesin-mediated loops via chromatin interaction analysis by paired-end tag sequencing (ChIA-PET). We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and resampling techniques. Our results demonstrate that genomic distance between boundary regions and genomic annotations coupled with random under-sampling results in the best model performance. We show that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) is sufficient for accurate boundary predictions. *preciseTAD* is implemented as an R package (version 1.1.1; Bioconductor version 3.12) and provides the pre-trained models, via an ExperimentHub R package *preciseTADhub* (version 0.99.8; Bioconductor version 3.13), which allows for precise prediction of boundaries for any cell line for which the genomic annotations are available.
<!--%%%%-->

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (<!--Additional file 16: Table S5--> <!--%%%-->Additional file 20: Table S5<!--%%%-->). Our aim was to have a balanced metric sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs across different thresholds. However, it doesn't emphasize one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be susceptible to different values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance, preventing us from investigating its effect, and also omits from its calculation TN values. All of these are important considerations to make when choosing a performance metric. For these reasons, we opted to report balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data resampling.

@@

The DBSCAN algorithm has two parameters, *MinPts* and *eps* ($\epsilon$). The *MinPts* parameter was set to the recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. To decide on the optimal value of $t$ and $\epsilon$ in *preciseTAD*, we considered the normalized enrichment $NE$ of flanked predicted boundaries, defined as

@@

**Figure 5. *preciseTAD* predicts more biologically meaningful TAD boundaries** (A) The location of Arrowhead-called TAD boundaries (blue) vs. *preciseTAD*-predicted TAD boundaries (green) on GM12878 data (chr14:50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) A zoomed-in portion of the genome shows the preciseTAD boundary region (PTBR, highlighted yellow), a cluster of bases with high probability of being a boundary, and the corresponding signal profiles of CTCF, RAD21, SMC3, and ZNF143.

>>>>

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of the 3-dimensional (3D) structure of the human genome such as chromatin loops [@rao20143d], Topologically Associating Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At the kilobase scale, chromatin loops connect gene promoters with distal enhancers, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. At the megabase scale, TADs represent regions on the linear genome that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation [@phillips2013architectural]. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Thus, identifying the precise location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify genomic coordinates of TADs (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised]. However, a key limitation of them is that they depend on Hi-C data *resolution*. Resolution refers to the size of genomic regions (bins) used to segment the linear genome [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which leads to increased uncertainty in the location of boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. While these low-resolution Hi-C datasets can be used to probe higher-order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TAD and loop boundaries [@fudenberg2016formation]. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth [@schmitt2016genome], obtaining high-resolution Hi-C data and boundaries remains difficult.

@@

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin state (BroadHMM), histone modification (HM), and transcription factor binding site (TFBS) data to predict both low-resolution TAD and chromatin-loop boundaries. We introduce a systematic pipeline for building the optimal boundary region prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites improve prediction performance, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and partitioning around medoids (PAM) to detect genome annotation-guided boundary regions and points at a base-level resolution. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD*-predicted boundaries are more conserved across cell lines, and the model trained in one cell line can accurately predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package and supplementary *preciseTADhub* ExperimentHub package are freely available on Bioconductor.

@@

<!--%%%% revised paragraph-->
We developed a machine learning (ML) framework for determining the optimal set of data level characteristics to predict boundary regions of Topologically Associating Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. For TAD prediction, we derived cell line specific boundaries at 5-100 kb resolutions using *Arrowhead* [@durand2016juicer] (Additional file 1: Arrowhead Script, Additional file 2: Table S1). For chromatin loop prediction, we downloaded loops derived by *Peakachu* [@salameh2019supervised] at 10 kb (Additional file 2: Table S1). Our models were trained on cell line-specific functional genomic annotation data (ChIP-seq peak regions) publicly available through ENCODE [@encode2004encode; @davis2018encyclopedia] (Additional file 3: Table S2). We used ChIA-PET derived loop boundaries [@grubert2020landscape] (Additional file 2: Table S1), referred hereafter as *Grubert* boundaries, as the gold standard when validating our models and extracting performance metrics (See Methods). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base-level resolution from cell line-specific genome annotation data.
<!--%%%-->

@@

Boundary regions were defined as genomic bins containing a called boundary ($Y = 1$), while non-boundary regions were defined as bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Additional file 4: Table S3). The number of non-boundary regions highly outnumbered boundary regions. Such a disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address the class imbalance, we evaluated the effect of three resampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority class (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

@@

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracy than the peak signal strength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions, with results consistent for K562 (Additional file 6: Figure S2A). Furthermore, 5 kb resolution genomic bins led to the optimal prediction for TAD boundary regions on both cell lines. Random forest models built on Peakachu-defined loop boundary regions yielded optimal prediction performance when using a distance-type predictor space, with RUS marginally outperforming SMOTE, for both cell lines (Additional file 6: Figure S2B, S2C). Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and chromatin loop boundary prediction.

@@

We sought to further optimize our boundary region prediction models by implementing recursive feature elimination to avoid overfitting and selecting only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Additional file 7: Figure S3B). However, given that we trained our models on chromosome-specific data, the most significant annotations varied for each holdout chromosome. To determine transcription factors most important for boundary prediction across all chromosomes, we clustered the predictive importance (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found four transcription factors, CTCF, RAD21, SMC3, and ZNF143, being consistently predictive of TAD and loop boundaries (Figure 4C; Additional file 7: Figure S3C). We optimized our model by only considering these top four TFBS when building the random forest model, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the known molecular drivers of the loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa], suggesting that TAD and loop formation are carried out by similar mechanisms.

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line using the model pre-trained on data from another cell line (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic annotation data from the GM12878 cell line resulted in an average AUC=<!--0.792--> <!--%%%--> 0.840<!--%%%--> (Figure 4D). Interestingly, when training on the K562 cell line and testing on GM12878, the average AUC increased slightly to <!--0.795--> <!--%%%--> 0.899<!--%%%-->. Likewise, the average performance of models trained using Peakachu boundaries and genomic annotation data from the GM12878 cell line was comparable to models trained on K562-specific Peakachu boundaries and genomic annotations (Avg. AUC=<!--0.881--> <!--%%%--> 0.923<!--%%%--> and <!--0.874--> <!--%%%--> 0.876<!--%%%-->, respectively (Figure 4D)). These results were consistent when comparing training/testing strategies on K562 with training on GM12878 and testing on K562 data (Additional file 7: Figure S3D). In both instances the average ROC curves were found to be within 1 standard deviation of each other, suggesting that a model trained on data from one cell line performs well when using the data from another cell line, allowing for the opportunity to predict boundaries for cell lines that do not currently have Hi-C data available.

@@

When trained using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb, respectively, the *preciseTAD* model predicted a total of 12,258 TAD and 15,707 chromatin loop boundaries in GM12878, as well as 9,603 TAD and 11,154 chromatin loop boundaries in K562 cell line (Additional file 9: Table S4). We reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Additional file 4: Table S3). This can be attributed to Arrowhead's inflation of called TADs at 5 kb, that, when visualized, often do not correspond to domains enriched in internal interactions (Figure 5A) and signal of known drivers of domain boundaries (Figure 5B). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Additional file 4: Table S3). This can be attributed to Peakachu's use of only CTCF sites to call boundaries, while *preciseTAD* leverages four known drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant than those detected from Hi-C data alone.

@@

Next, we evaluated the biological significance of *preciseTAD* boundary points. The signal of four known molecular drivers of 3D chromatin (CTCF, RAD21, SMC3, and ZNF143) colocalized more frequently around *preciseTAD*-predicted boundary points, as compared to Arrowhead-called TAD and Peakachu loop boundaries, respectively (<!--Figure 5C, 5D; Additional file 10: Figure S5A, S5B--> <!--%%%-->Figure 6A, 6B; Additional file 10: Figure S5A, S5B <!--%%%-->). <!--%%%-->Additionally, we found that when trained on cohesin-mediated chromatin loops, *preciseTAD*-predicted boundaries showed similar levels of colocalization to the set of ChIA-PET derived loop boundaries (Figure 6C; Additional file 10: Figure S5C).<!--%%%--> Surprised by the poor signal distribution around Arrowhead and Peakachu boundaries, we compared signals centered on boundaries called by Arrowhead, Peakachu, <!--%%%-->Grubert,<!--%%%--> and our SpectralTAD [@Cresswell:2020ab] TAD caller. We confirmed the poor signal distribution around Arrowhead-, Peakachu-, and SpectralTAD-called boundaries,in contrast to the well-performing Grubert-defined boundaries (<!--Additional file 10: Figure S5C, S5D--> <!--%%%-->Additional File 11: Figure S6<!--%%%-->). Signal enrichment heatmaps confirmed that *preciseTAD*-predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone<!--%%%-->, while showing similar levels of enrichment when compared to Grubert<!--%%%--> (<!--Additional file 11-12: Figure S6-S7--> <!--%%%-->Additional File 12-14: Figure S7-S9<!--%%%-->). <!--%%%-->While *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBS compared to either Arrowhead and Peakachu boundaries alone, Grubert boundaries were found to be the most spatially proximal (Wilcoxon p-value < 0.001; Figure 7A; Additional file 15: Figure S10). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation when compared to a novel TAD-caller or loop prediction algorithm.<!--%%%-->

@@

Guided by the high predictive performance when training and testing on different cell lines (Figure 4D), we opted to evaluate whether models trained using Arrowhead/Peakachu<!--%%%-->/Grubert<!--%%%--> ground truth data using *preciseTAD* in one cell line could be leveraged to predict boundaries using annotation data from another cell line. We evaluated two scenarios: 1) training on GM12878 and predicting boundaries on GM12878 (GM on GM) vs. training on K562 and predicting on GM12878 (K on GM), and 2) training on K562 and predicting boundaries on K562 (K on K) vs. training on GM12878 and predicting boundaries on K562 (GM on K). Using Arrowhead-trained models, 76% (J=0.701) and 81% (J=0.751) of predicted boundaries overlapped in both cross-cell-line prediction scenarios (<!--Additional file 15: Figure S10A, S10B--> <!--%%%-->Additional file 17: Figure S12<!--%%%-->). When using Peakachu-trained models, we observed 85% (J=0.705) and 88% (J=0.759) overlap (<!--Additional file 15: Figure S10C, S10D--> <!--%%%-->Additional file 18: Figure S13<!--%%%-->). <!--%%%-->Lastly, when using Grubert-trained models, we observed 86% (J=0.698) and 80% (J=0.0.660) overlap (Additional file 19: Figure S14).<!--%%%--> Furthermore, boundaries predicted on unseen annotation data exhibited a similar level of enrichment for CTCF, RAD21, SMC3, and ZNF143 as did those trained and predicted on the same cell line (Additional files 17-19: Figure S12-S14). These results indicate that *preciseTAD* pre-trained models can be successfully used to accurately predict both TAD and loop boundaries for cell lines lacking Hi-C data but for which genome annotation data is available.

@@

<!--%%%%revised paragraph-->
We present *preciseTAD*, a machine learning approach for precise prediction of TAD and chromatin loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on low-resolution domain boundaries obtained from domain calling tools, and high-resolution genomic annotations as the predictor space. *preciseTAD* predicts the probability of each base being a boundary, and identifies the precise location of boundary regions and the most likely boundary points. *preciseTAD* was benchmarked against multiple boundary calling tools including Arrowhead [@durand2016juicebox], an established TAD-caller, Peakachu [@salameh2019supervised], a recently published algorithm for predicting chromatin loops, and Grubert [@grubert2020landscape], a method for mapping cohesin-mediated loops via chromatin interaction analysis by paired-end tag sequencing (ChIA-PET). We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and resampling techniques. Our results demonstrate that genomic distance between boundary regions and genomic annotations coupled with random under-sampling results in the best model performance. We show that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) is sufficient for accurate boundary predictions. *preciseTAD* is implemented as an R package (version 1.1.1; Bioconductor version 3.12), while pre-trained boundary region prediction models are provided via an ExperimentHub R package *preciseTADhub* (version 0.99.8; Bioconductor version 3.13). These open source tools allow for precise prediction of boundaries for any cell line for which the genomic annotations are available.
<!--%%%%-->

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (<!--Additional file 16: Table S5--> <!--%%%-->Additional file 20: Table S5<!--%%%-->). Our aim was to have a balanced metric sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs across different thresholds. However, it doesn't emphasize one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be susceptible to different values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance, preventing us from investigating its effect, and also omits TN values from its calculation. All of these are important considerations to make when choosing a performance metric. For these reasons, we opted to report balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data resampling.

@@

The DBSCAN algorithm has two parameters, *MinPts* and *eps* ($\epsilon$). The *MinPts* parameter was set to the recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. To decide on the optimal value of $t$ and $\epsilon$ in *preciseTAD*, we considered the normalized enrichment ($NE$) of flanked predicted boundaries, defined as

@@

**Figure 5. *preciseTAD* predicts more biologically meaningful TAD boundaries** (A) The location of Arrowhead-called TAD boundaries (blue) vs. *preciseTAD*-predicted TAD boundaries (green) on GM12878 data (chr14:50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) A zoomed-in portion of the genome shows the preciseTAD boundary region (PTBR, highlighted yellow), a cluster of bases with high probability of being a boundary, and the corresponding binding signal of CTCF, RAD21, SMC3, and ZNF143.

**Figure 6. *preciseTAD* boundaries are more enriched for known molecular drivers of 3D chromatin.** Signal profile plots illustrating enrichment strength of CTCF, RAD21, SMC3, and ZNF143 sites around *preciseTAD*-predicted boundaries (green) compared to (A) Arrowhead-called boundaries (blue), (B) Peakachu loop boundaries (red), and (C) Grubert loop boundaries (purple). Boundaries involving Arrowhead and Grubert were flanked by 5 kb, while Peakachu-related boundaries were flanked by 10 kb.
